---
title: "Day 1 as a Business Analyst: Understanding the Financial Landscape"
author: "Jonathan Elkins"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    code_folding: show
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 10,
  fig.height = 6,
  fig.align  = "center"
)

set.seed(42)


```

```{r, libraries}

library(tidyverse)
library(scales)
library(lubridate)
library(ggthemes)
library(knitr)
library(kableExtra)
library(patchwork)

```

# Setting the Scene

For this analysis I am stepping into the role of a newly hired business analyst
at a mid-market SaaS company. The business sells software on a subscription
basis, meaning customers pay a recurring monthly fee in exchange for access to
the product. The company has been operating for a few years, has an established
customer base, and is in a growth phase — adding new customers every month while
managing the costs of building the product, selling it, and supporting the people
who use it.

The company has five departments. Sales owns new customer acquisition. Marketing
generates pipeline and manages the brand. Engineering builds and maintains the
product. G&A covers finance, legal, HR, and the other functions that keep the
business running. Customer Success manages relationships with existing customers
and is responsible for keeping them from canceling their subscriptions.

This matters because every number we look at today connects back to one of those
five departments and the decisions the people in them are making. Revenue does
not just happen. It is the result of salespeople closing deals, marketers
generating leads, engineers shipping features, and customer success teams keeping
accounts healthy. The data tells us where those efforts are working and where
they are not.

The analysis covers 24 months of P&L data from January 2023 through December
2024, 12 months of budget vs. actual data for 2024, a current headcount snapshot,
and 24 months of customer acquisition and retention metrics. By the end I want
a clear picture of the financial health of this business and a short list of
things I need to understand better before I start building anything.

My job right now is not to have answers. It is to ask the right questions. That
means working through four lenses:

1. **Revenue & Margin** — Are we growing? Is that growth profitable?
2. **Budget vs. Actual** — Where are we spending differently than planned?
3. **Headcount & Labor** — Are we staffed efficiently?
4. **Unit Economics** — What does it actually cost to serve a customer?

I am going to walk through each one out loud, showing my code, my reasoning,
and the questions each chart raises. By the end I want a short list of things
to follow up on in my first week.

> **Note on data:** The data here is synthetic but built to mirror realistic
> patterns. Seasonality, departmental cost structures, headcount ratios, margin
> profiles you would actually encounter. The code is fully reproducible and the
> patterns are real even if the numbers are not.

# Generating the Data

Before I touch a single chart I need to build the dataset. In a real job this
is where I would open a ticket with the data engineering team, query a warehouse
like Snowflake or BigQuery, or ask a financial analyst to pull the latest close
from NetSuite. The data would be messy, have gaps, and require cleaning before
it was usable.

Here I am generating everything synthetically in R. That means I am writing code
that produces fake but realistic numbers — the kind of numbers a real business
in this revenue range and growth stage might actually produce. Seasonality is
built in. Noise is built in. The relationships between variables are economically
logical. The goal is not to simulate a specific company but to create a dataset
that behaves like one so the analysis that follows reflects real analytical
thinking rather than textbook examples.

Every dataset is printed after it is built so you can see exactly what structure
I am working with before any analysis begins. If you handed me a real export
from your finance system this is the first thing I would do — look at the shape
of the data before I build anything on top of it.

```{r generate-data}

months <- seq.Date(from = as.Date("2023-01-01"), by = "month", length.out = 24)

monthly_pl <- tibble(
  month = months,
  revenue = 650000 * (1 + 0.015 * (1:24)) *
    (1 + 0.08 * sin((month(months) - 3) * pi / 6)) +
    rnorm(24, 0, 15000),
  cogs = revenue * (0.52 - 0.003 * (1:24) / 24) + rnorm(24, 0, 8000),
  opex_salaries = 180000 + 2500 * (1:24) + rnorm(24, 0, 5000), 
  opex_marketing = 45000 + 800 * (1:24) + rnorm(24, 0, 3000),
  opex_tech      = 28000 + 400 * (1:24) + rnorm(24, 0, 1500),
  opex_facilities = 32000 + rnorm(24, 0, 1000),
  opex_other      = 18000 + 300 * (1:24) + rnorm(24, 0, 2000)
) %>% 
  mutate(
    gross_profit = revenue - cogs,
    gross_margin = gross_profit / revenue,
    total_opex = opex_salaries + opex_marketing + opex_tech + opex_facilities + opex_other,
    ebitda = gross_profit - total_opex,
    ebitda_margin = ebitda / revenue,
    year = year(month),
    month_label = format(month, "%b %Y")
  )

```

This is the core P&L table. Each row is one month. Revenue and COGS are the
top two lines — the difference between them is gross profit. Below that are
five operating expense categories that together make up total OpEx. The derived
columns at the end — gross profit, gross margin, EBITDA, EBITDA margin — are
calculated from those raw inputs. Nothing in the derived columns is assumed.
Everything flows from the numbers above it.

```{r budget-data}

current_years_months <- months[year(months) == 2024]

budget_vs_actual <- tibble(
  month = rep(current_years_months, 5),
  department = rep(c("Sales", "Marketing", "Engineering", "G&A", "Customer Success"), each = length(current_years_months))
) %>% 
  mutate(
    budget = case_when(
      department == "Sales"            ~ 95000  + 1000  * as.integer(factor(month)), 
      department == "Marketing"        ~ 48000  + 500   * as.integer(factor(month)), 
      department == "Engineering"      ~ 120000 + 800   * as.integer(factor(month)),  
      department == "G&A"              ~ 42000  + 200   * as.integer(factor(month)), 
      department == "Customer Success" ~ 35000  + 300   * as.integer(factor(month)) 
    ),
    variance_pct = case_when(
      department == "Sales"            ~ rnorm(n(), 0.06,  0.04), 
      department == "Marketing"        ~ rnorm(n(), 0.12,  0.05), 
      department == "Engineering"      ~ rnorm(n(), -0.03, 0.03),  
      department == "G&A"              ~ rnorm(n(), 0.02,  0.02),
      department == "Customer Success" ~ rnorm(n(), -0.05, 0.03)
    ),
    actual = budget * (1 + variance_pct),
    variance = actual - budget
  )
```

This table covers 2024 only because budget vs. actual is always a current year
conversation. Each row is one department in one month. Budget is what we planned
to spend. Actual is what we spent. Variance is the dollar difference and variance
percent is that gap as a proportion of budget. A positive variance means we
spent more than planned. A negative variance means we spent less.

```{r headcount-data}
headcount <- tibble(
  department = c("Sales", "Marketing", "Engineering", "G&A", "Customer Success"),
  headcount = c(22, 8, 31, 9, 14),
  avg_salary = c(82000, 78000, 115000, 72000, 68000),
  avg_tenure_yrs = c(2.1, 1.8, 3.4, 4.2, 2.7),
  open_reqs      = c(3, 1, 5, 0, 2),
  attrition_rate = c(0.18, 0.22, 0.12, 0.08, 0.25)
) %>% 
  mutate(
    total_labor_cost = headcount * avg_salary,
    fully_loaded = total_labor_cost * 1.25
  )

```

Headcount is a snapshot not a time series. Five departments, eight columns. The
two derived columns at the end — total labor cost and fully loaded — are
calculated here. Fully loaded multiplies base salary cost by 1.25 to account
for benefits, payroll taxes, and employer contributions. That multiplier is a
standard finance rule of thumb. In a real job I would get the exact figure from
HR or the finance team but 1.2 to 1.3 is the typical range.

```{r unit-econ-data}

unit_econ <- tibble(
 month             = months,
  new_customers     = round(120 + 4 * (1:24) + rnorm(24, 0, 12)),
  churned_customers = round(40 + 1.5 * (1:24) + rnorm(24, 0, 4)),
  sales_marketing_spend = monthly_pl$opex_salaries * 0.35 +
                          monthly_pl$opex_marketing
) %>% 
  mutate(
    total_customers   = cumsum(new_customers - churned_customers) + 800,
    cac               = sales_marketing_spend / pmax(new_customers, 1),
    arpu              = (monthly_pl$revenue / total_customers) * 0.25,
    churn_rate        = churned_customers / total_customers,
    ltv               = arpu / pmax(churn_rate, 0.001) * 0.70,
    ltv_cac_ratio     = ltv / cac,
    cogs_per_customer = monthly_pl$cogs / total_customers
  )

```

Unit economics tracks how the business acquires and retains customers over time.
New customers and churned customers drive the running total of active accounts.
CAC is how much we spend to win each new customer. ARPU is how much each
customer generates per month on average. Churn rate is the percentage of
customers we lose each month. LTV is the estimated lifetime value of a customer
given their ARPU and how long they typically stay. The LTV to CAC ratio is the
single number that tells you whether the growth engine is healthy or not. A
ratio below 3x is a warning sign. Above 5x might mean you are underinvesting
in growth.

```{r data-check}

glimpse(monthly_pl)
glimpse(budget_vs_actual)
glimpse(headcount)
glimpse(unit_econ)

```

```{r theme-setup}

theme_ba <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 14,
                                margin = margin(b = 8)),
      plot.subtitle = element_text(color = "grey40", size = 11,
                                   margin = margin(b = 12)),
      plot.caption = element_text(color = "grey55", size = 9,
                                  margin = margin(t = 10)),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey92"),
      axis.text        = element_text(color = "grey40"),
      legend.position  = "bottom",
      strip.text       = element_text(face = "bold", size = 11)
    )
}

palette_dept <- c(
  "Sales"            = "#2E86AB",
  "Marketing"        = "#A23B72",
  "Engineering"      = "#F18F01",
  "G&A"              = "#C73E1D",
  "Customer Success" = "#3B1F2B"
)

```

# Revenue & Margin Analysis

The first thing I want to see is the revenue trend. Not just the number but the
shape. Is growth accelerating or decelerating? Is it smooth or lumpy? Does it
show seasonality I need to account for in budget models? A single number tells
you where you are. A trend tells you where you are going.

```{r revenue-trend} 

monthly_pl %>% 
  ggplot(aes(x = month, y = revenue)) +
  geom_area(fill = "#2E86AB", alpha = 0.15) +
  geom_line(color = "#2E86AB", linewidth = 1.2) +
  geom_point(color = "#2E86AB", size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = "grey40",
              linetype = "dashed", linewidth = 0.8) +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K"))+
  scale_x_date(date_labels = "%b `%y", date_breaks = "3 months") +
  labs(
    title = "Monthly Revenue - 24 Month Trend",
    subtitle = "Dashed line shows linear trend | Revenue softer in Q4 each year",
    x = NULL,
    y = "Revenue",
    caption = "Source: Internal P&L"
  ) + 
  theme_ba()

```

Revenue is growing steadily across the full 24 month period and the linear trend
line confirms the direction is consistently upward. The area fill makes the
growth shape easy to read at a glance. What I notice immediately is the softness
in Q4 of both years — revenue dips slightly in the fall before recovering. That
is not a crisis but it is a pattern worth flagging. If the business is building
budgets that assume linear growth they are likely overestimating Q4 and
underestimating the recovery in Q1 and Q2. I would want to understand whether
this is demand driven, a timing issue in how contracts are signed, or something
in how revenue is recognized.

Revenue growth is encouraging but growth that erodes margin is not growth worth
celebrating. The top line going up means nothing if we are spending more to
deliver the product than we are taking in. Gross margin tells me whether the
core economics of the business are healthy — before we even consider what it
costs to run the company.

```{r gross-margin}

monthly_pl %>% 
  ggplot(aes(x = month, y = gross_margin)) +
  geom_line(color = "#27AE60", linewidth = 1.2) +
  geom_point(color = "#27AE60", size = 2.5) +
  geom_hline(yintercept = mean(monthly_pl$gross_margin),
             linetype = "dashed", color = "grey50") +
  annotate("text",
           x     = min(monthly_pl$month) + 30,
           y     = mean(monthly_pl$gross_margin) + 0.004,
           label = paste0("Avg: ", percent(mean(monthly_pl$gross_margin),
                                           accuracy = 0.1)),
           color = "grey40", size = 3.5) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1),
                     limits = c(0.44, 0.52)) +
  scale_x_date(date_labels = "%b '%y", date_breaks = "3 months") +
  labs(
    title = "Gross Margin % - 24 Month Trend",
    subtitle = "Volatile in 2023, stabilizing above average through 2024",
    x = NULL,
    y = "Gross Margin %",
    caption = "Gross Margin = (Revenue - COGS) / Revenue"
  ) +
  theme_ba()

```

Gross margin has averaged 48.4% across the period and has been stabilizing
above that average through 2024. The volatility in 2023 is worth noting —
the spread between the best and worst months was nearly 4 percentage points.
That kind of swing usually means COGS has a variable component that is not
fully under control, whether that is hosting costs, third party services, or
delivery costs that fluctuate with volume. By 2024 the band has tightened
which suggests the business has gotten better at managing those costs. That
is a good sign but I would want to understand what changed.

Gross margin tells me about the core business. EBITDA tells me about the whole
operation. After we pay our people, run our marketing, keep the lights on, and
cover everything else it takes to operate — what is actually left? This is where
I find out if the revenue growth we just saw is translating to the bottom line
or getting absorbed by the cost of running the company.

```{r ebitda-margin}

monthly_pl %>% 
  ggplot(aes(x = month, y = ebitda_margin,
             fill = ebitda_margin >= 0)) +
  geom_col(alpha = 0.85) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_x_date(date_labels = "%b '%y", date_breaks = "3 months") +
  scale_fill_manual(values = c("TRUE" = "#27AE60", "FALSE" = "#E74C3C"),
                    guide = "none") +
  labs(
    title    = "EBITDA Margin % — Monthly",
    subtitle = "Green = profitable months | Red = operating loss",
    x = NULL,
    y = "EBITDA Margin %",
    caption  = "EBITDA = Gross Profit minus all Operating Expenses"
  ) +
  theme_ba()

```

This is the most important chart in the section. The business was losing money
at the operating level in its earliest months — EBITDA was negative in January
and February of 2023. That is not unusual for a company investing heavily in
headcount and infrastructure ahead of revenue growth. What matters is the
direction. By mid 2023 the business had turned the corner and has been mostly
EBITDA positive since, with a few exceptions that coincide with the revenue dips
we saw in the first chart. The margins are thin — most positive months are in
the 2 to 7 percent range — which means there is not a lot of cushion if revenue
misses or costs spike unexpectedly. That is the watch item coming out of this
section.

# Budget vs. Actual Variance Analysis

This is often where a new analyst can add immediate value. The budget was set
months ago by people making assumptions about the future. Actuals are what
actually happened. The gap between them is where the interesting questions live.
My job here is not to assign blame but to find the patterns and ask why.

A budget variance table is one of the first things I would ask for on day one.
It tells me immediately which departments are running hot, which are running
cold, and whether the business is tracking to its financial plan. Before I look
at a single chart I want to see the numbers in a table so I can orient myself
to the scale of the gaps.

```{r bva-table}

bva_summary <- budget_vs_actual %>% 
  group_by(department) %>% 
  summarize(
    total_budget = sum(budget),
    total_actual = sum(actual),
    total_variance = sum(variance),
    variance_pct = total_variance / total_budget,
    .groups = 'drop'
  ) %>% 
  arrange(desc(variance_pct)) 

bva_summary %>% 
  mutate(
    total_budget = dollar(total_budget),
    total_actual = dollar(total_actual),
    total_variance = dollar(total_variance),
    variance_pct = percent(variance_pct, accuracy = 0.1)
  ) %>% 
  kable(
    col.names = c(
      "Department", "Budget", "Actual",
      "Variance ($)", "Variance (%)"
    ),
    align = c("l", "r", "r", "r", "r")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )


```

Three departments overspent and two underspent. Marketing is the biggest concern
at 13.2% over budget — that is not a rounding error, that is a pattern. Sales
is also over but commission structures are variable by nature so some overage
is expected when the team is performing. G&A at 2.5% over is noise. The more
interesting story is on the underspend side. Engineering is 3.6% under budget
and has 5 open requisitions sitting unfilled. That is not savings — that is
deferred hiring cost that will hit the moment those roles get filled. Customer
Success is 3.1% under which sounds fine until you remember their attrition rate.
Under budget and understaffed is not a good combination in a department
responsible for keeping customers.

The bar chart below makes the relative size of these gaps easier to read at a
glance.

```{r bva-chart}

bva_summary %>%
  ggplot(aes(x = reorder(department, variance_pct),
             y = variance_pct,
             fill = variance_pct > 0)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = paste0(round(variance_pct * 100, 1), "%"),
                hjust = ifelse(variance_pct > 0, -0.1, 1.1)),
            size = 4, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(labels = percent_format(),
                     limits = c(-0.1, 0.15)) +
  scale_fill_manual(values = c("TRUE" = "#E74C3C", "FALSE" = "#27AE60"),
                    guide = "none") +
  labs(
    title    = "Budget vs. Actual — YTD Variance by Department",
    subtitle = "Red = overspend | Green = underspend",
    x = NULL,
    y = "Variance %",
    caption  = "2024 YTD | Positive = actual exceeded budget"
  ) +
  theme_ba()

```

The visual confirms what the table showed but makes the relative scale
immediately clear. Marketing's overspend dwarfs everything else on a percentage
basis. Engineering and Customer Success are the only departments running under
and for reasons that deserve scrutiny rather than celebration.

A single YTD number can hide a lot. A department that blew its budget in January
but has been tightening every month since is a very different conversation than
one that keeps drifting further over. I want to see the monthly shape of the
variance for the three most interesting departments so I can understand whether
these are improving problems or worsening ones.

```{r bva-monthly}
budget_vs_actual %>% 
  filter(department %in% c("Marketing", "Sales", "Customer Success")) %>% 
  mutate(department = factor(department,
                             levels = c("Marketing", "Sales",
                                        "Customer Success"))) %>% 
  ggplot(aes(x = month, y = variance_pct, color = department)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  facet_wrap(~department, ncol = 1, scales = "free_y") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_date(date_labels = "%b '%y") +
  scale_color_manual(values = palette_dept, guide = "none") +
  labs(
    title = "Monthly Variance % - Overspending Departments",
    subtitle = "Positive = spending above budget | Negative = under budget",
    x = NULL,
    y = "Variance %"
  ) +
  theme_ba()
```

Marketing is consistently over budget every single month with no meaningful
improvement. That is not a one-time event — it is a structural overspend that
someone either approved and did not communicate or has not noticed yet. Sales
is volatile, swinging above and below budget month to month, which is consistent
with a variable commission structure tied to deal timing. Customer Success has
been running under all year but is trending toward breakeven by December. Cross
referenced with their attrition rate that trend is worth watching — if they
start backfilling roles aggressively that underspend will flip quickly.

# Headcount & Labor Cost Analysis

Labor is almost always the largest expense line in a business, often 60 to 70
percent of total OpEx. Understanding who we have, what they cost, and whether
we are retaining them is non-negotiable on day one. The headcount data also
connects directly back to what we just saw in the budget variance section.
Departments that are overspending or understaffed do not exist in isolation —
they show up here too.

The table below is the starting point. Before I build any charts I want to see
the raw numbers — headcount, salary, tenure, open roles, and attrition — all
in one place.

```{r headcount-table}
headcount_display <- headcount %>% 
  arrange(desc(fully_loaded)) %>% 
  mutate(
    avg_salary       = dollar(avg_salary),
    total_labor_cost = dollar(total_labor_cost),
    fully_loaded     = dollar(fully_loaded),
    attrition_rate   = percent(attrition_rate, accuracy = 1)
  )

headcount_display %>%
  kable(
    col.names = c("Department", "Headcount", "Avg Salary",
                  "Avg Tenure (Yrs)", "Open Reqs", "Attrition %",
                  "Total Labor", "Fully Loaded"),
    align = c("l", "r", "r", "r", "r", "r", "r", "r")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE)
```

A few things jump out immediately. Engineering is by far the largest cost center
at $4.4 million fully loaded annually and has the most open roles at 5. G&A has
the highest average tenure at 4.2 years which makes sense — finance, legal, and
HR tend to be stable functions. Customer Success has the lowest average salary
at $68,000 and the highest attrition at 25%. That combination is worth pausing
on. The people responsible for keeping customers happy are the least compensated
and the most likely to leave. The two charts below break this down further.

The first view compares fully loaded labor cost across departments. The second
looks at attrition rate with a threshold line at 18% — departments above that
line are losing people at a rate that creates real operational risk.

```{r headcount-charts}

p_hc1 <- headcount %>% 
  ggplot(aes(x = reorder(department, fully_loaded),
             y = fully_loaded,
             fill = department)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = dollar(fully_loaded)),
            hjust = -0.1, size = 3.5, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(labels = dollar_format(scale = 1e-6, suffix = "M"),
                     limits = c(0, max(headcount$fully_loaded) * 1.35)) +
  scale_fill_manual(values = palette_dept, guide = "none") +
  labs(
    title = "Fully Loaded Labor Cost",
    subtitle = "Base Salary * 1.25",
    x = NULL,
    y = "Annual Cost"
  ) +
  theme_ba()

p_hc2 <- headcount %>% 
  ggplot(aes(x = reorder(department, attrition_rate),
             y = attrition_rate,
             fill = attrition_rate >= 0.18)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = percent(attrition_rate, accuracy = 1)),
            hjust = -0.1, size = 3.5, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(labels = percent_format(),
                     limits = c(0, 0.32)) +
  scale_fill_manual(values = c("FALSE" = "#27AE60", "TRUE" = "#E74C3C"),
                    guide = "none") +
  labs(
    title = "Annual Attrition Rate",
    subtitle = "Red = above 18% threshold",
    x = NULL,
    y = "Attrition %"
  ) +
  theme_ba()

p_hc1 + p_hc2

```

Engineering dominates the cost chart which is expected given the combination of
headcount and salary. What stands out on the attrition chart is that three of
five departments are at or above the 18% threshold. That is not a one department
problem — it is a company wide retention issue that finance leadership needs to
be aware of.

The bar charts show each dimension separately. This next view combines them.
The goal is to find departments where high cost and high attrition overlap —
because that is where the business is taking the most financial risk. Every
person who leaves has to be replaced, and replacement costs typically run 50
to 100 percent of annual salary when you account for recruiting, onboarding,
and lost productivity.

```{r headcount-scatter}
headcount %>% 
  ggplot(aes(x = attrition_rate, y = avg_salary,
             size = headcount, color = department, label = department)) +
  geom_point(alpha = 0.75) +
  geom_text(vjust = -1.2, hjust = 0.5, size = 3.5, fontface = "bold") +
  geom_vline(xintercept = 0.18, linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = mean(headcount$avg_salary),
             linetype = "dashed", color = "grey50") +
  annotate("text", x = 0.185, y = 69000,
           label = "Attrition threshold",
           color = "grey50", size = 3, hjust = 0) +
  annotate("text", x = 0.09, 
           y = mean(headcount$avg_salary) + 3500,
           label = "Avg Salary",
           color = "grey50", size = 3) +
  scale_x_continuous(labels = percent_format(), limits = c(0.07, .28)) +
  scale_y_continuous(labels = dollar_format(), limits = c(NA, 125000)) +
  scale_color_manual(values = palette_dept, guide = "none") +
  scale_size(range = c(3,10), guide = "none") +
  labs(
    title = "Attrition Rate vs. Average Salary",
    subtitle = "Bubble size = headcount | Top right = expensive and high turnover",
    x = "Annual Attrition Rate",
    y = "Average Salary"
  ) +
  theme_ba()
  
```

Engineering sits in the top left — expensive but well retained. That is the
ideal position. G&A sits in the bottom left — lower cost and stable. Customer
Success sits in the bottom right — lower salary but the highest attrition in
the company. That is the danger quadrant. You are not paying a lot per person
but you are cycling through people constantly in a role where relationship
continuity directly affects whether customers stay or leave. Marketing is also
above the attrition threshold and above average salary which makes it the
second most concerning department from a retention standpoint. Sales sits right
on the threshold — worth watching but not yet a red flag on its own.

# Unit Economics

This is the section most junior analysts skip. Unit economics live at the
intersection of finance, product, and operations and understanding them tells
you whether the growth we saw in the revenue section is actually sustainable.
A business can grow revenue every month and still be destroying value if it
costs more to acquire and serve a customer than that customer is ever going
to return. The summary table below gives me a snapshot of where things stand
at the end of the period.

```{r unit-econ-latest, results = 'hide'}
latest <- unit_econ %>% 
  slice_tail(n = 1)

latest
```

```{r unit-econ-display}

unit_econ_display <- tibble(
  Metric = c(
    "Total Customers",
    "New Customers (Latest Month)",
    "Churned Customers (Latest Month)",
    "Customer Acquisition Cost",
    "Monthly ARPU",
    "Monthly Churn Rate",
    "Estimated LTV",
    "LTV:CAC Ratio"
  ),
  Value = c(
    comma(latest$total_customers),
    comma(latest$new_customers),
    comma(latest$churned_customers),
    dollar(latest$cac),
    dollar(latest$arpu),
    percent(latest$churn_rate, accuracy = 0.1),
    dollar(latest$ltv),
    round(latest$ltv_cac_ratio, 2)
  ),
  Interpretation = c(
    "Current book of business",
    "Gross new logos this month",
    "Accounts lost this month",
    "Sales and marketing spend per new customer",
    "Average revenue per user per month",
    "Percent of customers lost per month",
    "ARPU divided by churn rate times 70% margin",
    "Target is 3x or above for a healthy business"
  )
)

unit_econ_display %>%
  kable(align = c("l", "r", "l")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)

```

The LTV to CAC ratio at 2.56 is the number that concerns me most. It is below
the 3x threshold that most operators treat as the minimum for a healthy
acquisition engine. That means for every dollar we spend acquiring a customer
we are getting back roughly $2.56 in lifetime value. That is not catastrophic
but it is not comfortable either. The three charts below tell me how we got
here and where things are heading.

Customer acquisition cost tells me how efficiently the sales and marketing
machine is converting spend into new customers. If CAC is rising it means we
are working harder and spending more to win the same number of accounts. If it
is falling it means we are getting more efficient. The trend over 24 months
tells a more honest story than any single month.

```{r cac-chart}

unit_econ %>% 
  ggplot(aes(x = month, y = cac)) +
  geom_line(color = "#E74C3C", linewidth = 1.2) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.15,
              color = "#E74C3C", fill = "#E74C3C") +
  scale_y_continuous(labels = dollar_format()) +
  scale_x_date(date_labels = "%b '%y", date_breaks = "6 months") +
  labs(
    title = "Customer Acquisition Cost - 24 Month Trend",
    subtitle = "Declining CAC suggests improving sales efficiency over 24 months",
    x = NULL,
    y= "CAC ($)"
  ) +
  theme_ba()

```

CAC has been declining over the period which is an encouraging sign. The
business is getting more efficient at acquiring customers, likely through
improving sales processes, better marketing targeting, or both. The question
I would ask is whether this is intentional or whether we are simply picking
lower hanging fruit and deferring the harder prospects for later. Declining
CAC that is driven by cherry picking easy deals will reverse the moment those
deals run out.

CAC declining is good news but it does not tell the full story on its own.
What matters is whether the value of a customer over their lifetime justifies
what we spent to acquire them. That is the LTV to CAC ratio and it is the
single most important unit economics metric for understanding the health of
the growth engine.

```{r ltv-cac-chart}
unit_econ %>% 
  ggplot(aes(x=month, y = ltv_cac_ratio)) +
  geom_line(color = "#2E86AB", linewidth =1.2) +
  geom_point(color = "#2E86AB", size = 2.5) +
  geom_hline(yintercept = 3, linetype = 'dashed',
             color = "#27ae60", linewidth = 1) +
  annotate("text",
           x = min(unit_econ$month) + 30,
           y = 3.15,
           label = "Healthy threshold (3.0x)",
           color = "#27ae60", size = 3.5) +
  scale_x_date(date_labels = "%b '%y", date_breaks = "6 months") +
  scale_y_continuous(limits = c(2, NA)) +
    labs(
    title    = "LTV:CAC Ratio — 24 Month Trend",
    subtitle = "Below 3x means we are paying too much to acquire relative to customer value",
    x = NULL,
    y = "LTV:CAC"
  ) +
  theme_ba()
```

The ratio has been volatile across the period and is currently sitting at 2.56
which puts us below the 3x threshold. ARPU is declining as we bring on smaller
accounts and CAC is only declining modestly. The gap is closing in the wrong
direction. This is not a crisis today but it is the conversation I am walking
into on day two. If this trend continues for another two or three quarters the
business will need to either raise prices, reduce acquisition costs, or improve
retention to get the ratio back above 3x.

The last unit economics view is churn. Every percentage point of monthly churn
compounds. At 2% monthly you lose roughly 22% of your customer base annually.
At 3% that number climbs to 31%. Churn is the leaky bucket underneath all the
growth numbers we looked at in section three and it connects directly to the
Customer Success attrition problem we flagged in section five. The question is
not just where churn is today but where it has been and what direction it is
moving.

```{r churn-chart}
unit_econ %>%
  ggplot(aes(x = month, y = churn_rate)) +
  geom_line(color = "#A23B72", linewidth = 1.2) +
  geom_point(color = "#A23B72", size = 2.5) +
  geom_hline(yintercept = 0.02, linetype = "dashed",
             color = "grey50", linewidth = 0.8) +
  annotate("text",
           x     = min(unit_econ$month) + 30,
           y     = 0.0215,
           label = "2% monthly benchmark",
           color = "grey50", size = 3.5) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_x_date(date_labels = "%b '%y", date_breaks = "6 months") +
  labs(
    title    = "Monthly Customer Churn Rate — 24 Month Trend",
    subtitle = "Even small churn compounds significantly at scale",
    x = NULL,
    y = "Monthly Churn Rate"
  ) +
  theme_ba()
```

Churn started well above 5% in early 2023 which is a serious problem for any
subscription business. The encouraging news is the trend is consistently
downward and by late 2024 we are approaching the 2% monthly benchmark. That
kind of sustained improvement does not happen by accident. Something changed —
whether that was a CS process improvement, a product fix, or a change in the
customer mix. I would want to know exactly what drove it because whatever it
was we should be doing more of it. The concern is that with CS attrition at
25% the people who drove that improvement may not all still be here.

# Executive Summary

I have now looked at this business through four lenses. Here is what I would
walk into my manager's office to say at the end of day one. Not conclusions,
not recommendations — those come after I have had more conversations. But
these are the patterns I see, the questions they raise, and the things I want
to understand better before I start building anything.

```{r exec-summary}
tibble(
  Category = c(
    "Revenue Growth (YoY)",
    "Gross Margin (Latest)",
    "EBITDA Margin (Latest)",
    "Marketing Budget Variance (YTD)",
    "Engineering Budget Variance (YTD)",
    "Customer Success Attrition",
    "LTV:CAC Ratio (Latest)",
    "Monthly Churn Rate (Latest)"
  ),
  Value = c(
    percent(
      (monthly_pl %>% filter(year == 2024) %>% pull(revenue) %>% mean()) /
      (monthly_pl %>% filter(year == 2023) %>% pull(revenue) %>% mean()) - 1,
      accuracy = 0.1),
    percent(last(monthly_pl$gross_margin), accuracy = 0.1),
    percent(last(monthly_pl$ebitda_margin), accuracy = 0.1),
    percent(filter(bva_summary, department == "Marketing") %>% 
              pull(variance_pct), accuracy = 0.1),
    percent(filter(bva_summary, department == "Engineering") %>% 
              pull(variance_pct), accuracy = 0.1),
    percent(filter(headcount, department == "Customer Success") %>% 
              pull(attrition_rate), accuracy = 1),
    as.character(round(last(unit_econ$ltv_cac_ratio), 2)),
    percent(last(unit_econ$churn_rate), accuracy = 0.1)
  ),
Signal = c(
  "Strong",
  "Healthy", 
  "Watch",
  "Investigate",
  "Deferred hiring",
  "Investigate",
  "Below threshold",
  "Improving"
)
) %>%
  kable(align = c("l", "r", "c"), escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```
**What looks healthy:**

Revenue is growing at 13.5% year over year and gross margin is averaging above
48% across the period, though the latest month came in at 46.9% which is worth
monitoring. The core business model is working — we are delivering the product
at a reasonable cost and growing the top line. The churn improvement from above
5% to near 2% over two years is the most encouraging trend in the whole analysis
and suggests the product and customer success teams have made real progress on
retention.

**What I am watching:**

EBITDA margin is thin and has dipped negative in several months. The business
is not consistently profitable at the operating level and there is not much
cushion if revenue softens or costs spike. Engineering has 5 open roles and
is running under budget — that underspend is not savings, it is deferred cost
that will hit the moment hiring accelerates.

**What I am flagging for week one conversations:**

Marketing is 13.2% over budget every single month with no sign of the gap
closing. I want to understand whether that spend was approved and whether it
is producing results in the new customer numbers. Customer Success has a 25%
attrition rate, 2 open roles, and is running under budget simultaneously. That
is the data point that concerns me most. The churn improvement we just saw was
almost certainly driven by CS team efforts. If the people who built that
improvement keep leaving, churn will reverse. The LTV to CAC ratio at 2.56 is
already below the 3x threshold. A churn reversal would make that worse. These
three things — CS attrition, churn trajectory, and LTV to CAC — are connected
and mapping that connection is the first thing I am building on day two.