---
title: "US Household Spending Analysis"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    theme: default
    source_code: embed
---

```{r setup, include=FALSE}
library(readxl)
library(flexdashboard)
library(tidyverse)
library(plotly)
library(scales)
library(htmltools)
```

```{r}

data <- read_xlsx("data/cu-income-before-taxes-2024.xlsx", skip = 2)

brackets <- c("all", "lt15k", "15_30k", "30_40k", "40_50k", 
               "50_70k", "70_100k", "100_150k", "150_200k", "200k_plus")

colnames(data) <- c("item", brackets)

rows_of_interest <- c(6, 48, 53, 190, 392, 453, 479)
labels <- c("annual_income", "total_spend", "food", 
            "housing", "transport", "healthcare", "entertainment")

bls_data <- data %>%
  mutate(row = row_number()) %>%
  filter(row %in% rows_of_interest) %>%
  mutate(category = labels) %>%
  select(category, `15_30k`:`200k_plus`) %>%
  pivot_longer(-category, names_to = "bracket", values_to = "value") %>%
  mutate(value = as.numeric(value)) %>%
  pivot_wider(names_from = category, values_from = value) %>%
  mutate(
    savings = annual_income - total_spend,
    savings_rate = savings / annual_income,
    spend_ratio = total_spend / annual_income,
    income_bracket = factor(bracket,
                            levels = c("15_30k", "30_40k", "40_50k",
                                       "50_70k", "70_100k", "100_150k",
                                       "150_200k", "200k_plus"),
                            labels = c("$15-30k", "$30-40k", "$40-50k",
                                       "$50-70k", "$70-100k", "$100-150k",
                                       "$150-200k", "$200k+"))
  )

```

```{r}

spend_long <- bls_data %>% 
  select(income_bracket, food, housing, transport,
         healthcare, entertainment) %>% 
  pivot_longer(-income_bracket,
               names_to = "category",
               values_to = "amount") %>% 
  mutate(category = str_to_title(category))

```



```{r}

pal <- c("#00e5a0", "#7b5ea7", "#ff6b35", "#4ecdc4", "#ffe66d", "#a8e6cf")

```

```{css}
.card-title {
  font-size: 0.95rem !important;
  letter-spacing: 0.08em;
  text-transform: uppercase;
  opacity: 0.7;
}

.card-body {
  font-size: 2.8rem !important;
  font-weight: 700 !important;
}

.chart-stage table,
.chart-stage-flex table,
body table {
  width: 100% !important;
  font-size: 1.1rem !important;
  border-collapse: collapse !important;
}

.chart-stage th,
.chart-stage-flex th,
body th {
  background-color: #1a1a2e !important;
  color: #00e5a0 !important;
  padding: 0.85rem 1.2rem !important;
  text-align: left !important;
  letter-spacing: 0.08em !important;
  text-transform: uppercase !important;
  font-size: 0.9rem !important;
}

.chart-stage td,
.chart-stage-flex td,
body td {
  padding: 0.85rem 1.2rem !important;
  border-bottom: 1px solid #2a2a3a !important;
  vertical-align: top !important;
  line-height: 1.7 !important;
  font-size: 1.1rem !important;
}

tr:hover td {
  background-color: rgba(0, 229, 160, 0.05) !important;
}
```

Overview
=====================================

Row {data-height=140}
-------------------------------------

### Median Annual Income

```{r}

median_inc <- median(bls_data$annual_income)
valueBox(dollar(median_inc, accuracy = 1), color = "#1a1a2e")

```

### Median Savings Rate

```{r}

median_sar <- median(bls_data$savings_rate)
valueBox(percent(median_sar, accuracy = 0.1), color = "#1a1a2e")

```

### Median Spend Ratio

```{r}

median_spr <- median(bls_data$spend_ratio)
valueBox(percent(median_spr, accuracy = 0.1), color = "#1a1a2e")

```


### Brackets Running Deficit

```{r}

deficit_count <- sum(bls_data$savings < 0)
valueBox(deficit_count, color = "#1a1a2e")

```


Row {data-height=430}
-------------------------------------

### Spend Ratio Across Income Brackets

```{r}

plot_ly(bls_data,
        x = ~income_bracket,
        y = ~spend_ratio,
        type = "scatter",
        mode = "lines+markers",
        line = list(color = "#00e5a0", width = 3),
        markers = list(color = "#00e5a0", size = 8),
        hovertemplate =  "Bracket: %{x}<br>Spend Ratio: %{y:.1%}<extra></extra>") %>% 
  layout(
    shapes = list(list(
      type = "line", x0 = 0, x1 = 1, xref = "paper",
      y0 = 1, y1 = 1,
      line = list(color = "#ff6b35", dash = "dash", width = 2)
    )),
    xaxis = list(title = "Income Bracket", tickangle = -45),
    yaxis = list(title = "Spend Ratio", tickformat = ".0%")
  )


```

Row {data-height=430}
-------------------------------------

### Income vs. Average Annual Spend

```{r}

plot_ly(bls_data) %>% 
  add_trace(x = ~income_bracket, y = ~annual_income,
            type = "bar", name = "Income",
            marker = list(color = "#00e5a0"),
            hovertemplate = "Income: %{y:$,.0f}<extra></extra>") %>% 
  add_trace(x = ~income_bracket, y = ~total_spend,
            type = "bar", name = "Total Spend",
            marker = list(color = "#ff6b35"),
            hovertemplate = "Spend: %{y:$,.0f}<extra></extra>") %>% 
  layout(
    barmode = "group",
    xaxis = list(title = "Income Bracket", tickangle = -45),
    yaxis = list(title = "Amount ($)", tickformat = "$,.0f"),
    legend = list(orientation = "h", y = -0.3)
  )

```

Where Does Money Go?
=====================================

Row {data-height=500}
-------------------------------------

### Spending Category Breakdown by Bracket

```{r}

plot_ly(spend_long,
        x      = ~income_bracket,
        y      = ~amount,
        color  = ~category,
        colors = pal,
        type   = "bar",
        hovertemplate = "<b>%{x}</b><br>%{fullData.name}: %{y:$,.0f}<extra></extra>") %>%
  layout(
    barmode = "stack",
    xaxis = list(title = "Income Bracket", tickangle = -45),
    yaxis = list(title = "Avg Annual Spend ($)", tickformat = "$,.0f"),
    legend = list(orientation = "h", y = -0.25)
  )

```


Row {data-height=500}
-------------------------------------

### Category as % of Income
```{r}

pct_data <- bls_data %>%
  mutate(
    Housing       = housing / annual_income,
    Food          = food / annual_income,
    Transport     = transport / annual_income,
    Healthcare    = healthcare / annual_income,
    Entertainment = entertainment / annual_income
  ) %>%
  select(income_bracket, Housing, Food, Transport, Healthcare, Entertainment)

mat2 <- pct_data %>% select(-income_bracket) %>% as.matrix()
rownames(mat2) <- as.character(pct_data$income_bracket)

plot_ly(
  x = colnames(mat2),
  y = rownames(mat2),
  z = mat2,
  type = "heatmap",
  colorscale = list(c(0,"#1a1a2e"), c(0.5,"#7b5ea7"), c(1,"#00e5a0")),
  hovertemplate = "%{y}<br>%{x}: %{z:.1%}<extra></extra>"
) %>%
  layout(
    xaxis = list(title = "Category"),
    yaxis = list(title = "Income Bracket")
  )

```


### Overall Spending Share

```{r}

category_totals <- spend_long %>%
  group_by(category) %>%
  summarise(total = sum(amount), .groups = "drop")

plot_ly(category_totals,
        labels  = ~category,
        values  = ~total,
        type    = "pie",
        marker  = list(colors = pal,
                       line = list(color = "#0a0a0f", width = 1)),
        textinfo = "label+percent",
        hovertemplate = "<b>%{category}</b><br>Total: %{total:$,.0f}<extra></extra>") %>%
  layout(
    legend = list(orientation = "h", y = -0.2)
  )


```

The Housing Burden
=====================================

Row {data-height=500}
-------------------------------------

### Housing Cost by Income Bracket

```{r}

plot_ly(bls_data,
        x = ~income_bracket,
        y = ~housing,
        type = "bar",
        marker = list(color = ~ifelse(
          housing / annual_income > 0.30, "#ff6b35", "#00e5a0")),
        hovertemplate = "Bracket: %{x}<br>Housing: %{y:$,.0f}<extra></extra>") %>%
  layout(
    xaxis = list(title = "Income Bracket", tickangle = -45),
    yaxis = list(title = "Housing Spend ($)", tickformat = "$,.0f"),
    annotations = list(list(
      x = 0.02, y = 0.98, xref = "paper", yref = "paper",
      text = "🔴 Above 30% income = cost burdened",
      showarrow = FALSE, font = list(size = 11, color = "#ff6b35")
    ))
  )
```

### Housing as % of Income

```{r}

plot_ly(bls_data,
        x    = ~income_bracket,
        y    = ~housing / annual_income,
        type = "scatter",
        mode = "lines+markers",
        line   = list(color = "#00e5a0", width = 3),
        marker = list(color = "#00e5a0", size = 8),
        hovertemplate = "Bracket: %{x}<br>Housing: %{y:.1%}<extra></extra>") %>%
  layout(
    shapes = list(list(
      type = "line", x0 = 0, x1 = 1, xref = "paper",
      y0 = 0.30, y1 = 0.30,
      line = list(color = "#ff6b35", dash = "dash", width = 2)
    )),
    annotations = list(list(
      x = 0.01, y = 0.32, xref = "paper", yref = "y",
      text = "HUD Cost Burden Threshold (30%)",
      showarrow = FALSE, font = list(size = 11, color = "#ff6b35")
    )),
    xaxis = list(title = "Income Bracket", tickangle = -45),
    yaxis = list(title = "Housing as % of Income", tickformat = ".0%")
  )

```


Row {data-height=250}
-------------------------------------

### Housing Insight

```{r}

housing_burdened <- sum(bls_data$housing / bls_data$annual_income > 0.30)

findings_housing <- tibble(
  Finding = c(
    "Cost-burdened brackets",
    "Worst affected bracket",
    "Breakeven point",
    "Top earner housing spend",
    "HUD definition"
  ),
  Insight = c(
    paste0(housing_burdened, " of 8 income brackets spend over 30% of income on housing"),
    paste0("$15-30k households spend ",
           percent(bls_data$housing[bls_data$bracket == "15_30k"] /
                   bls_data$annual_income[bls_data$bracket == "15_30k"],
                   accuracy = 0.1), " of income on housing alone"),
    "$70-100k is where housing drops below the 30% cost burden threshold",
    paste0("$200k+ households spend ",
           dollar(bls_data$housing[bls_data$bracket == "200k_plus"], accuracy = 1),
           " on housing — 3x the lowest bracket in absolute terms"),
    "HUD defines 'cost burdened' as spending more than 30% of income on housing"
  )
)

knitr::kable(findings_housing, align = c("l", "l"))

```

Savings & Deficits
=====================================

Row {data-height=500}
-------------------------------------

### Savings by Income Bracket

```{r}

plot_ly(bls_data,
        x = ~income_bracket,
        y = ~savings,
        type = "bar",
        marker = list(color = ~ifelse(savings < 0, "#ff6b35", "#00e5a0")),
        hovertemplate = "Bracket: %{x}<br>Savings: %{y:$,.0f}<extra></extra>") %>% 
  layout(
    xaxis = list(title = "Income Bracket", tickangle = -45),
    yaxis = list(title = "Avg Savings ($)", tickformat = "$,.0f")
  )

```

### Savings Rate Across Brackets

```{r}

plot_ly(bls_data,
        x = ~income_bracket,
        y = ~savings_rate,
        type = "scatter",
        mode = "lines+markers",
        line = list(color = "#00e5a0", width = 3),
        marker = list(color = "#00e5a0", size = 8),
        hovertemplate = "Bracket: %{x}<br>Savings: %{y:.1%}<extra></extra>") %>%
  layout(
    shapes = list(list(
      type = "line", x0 = 0, x1 = 1, xref = "paper",
      y0 = 0, y1 = 0,
      line = list(color = "#ff6b35", dash = "dash", width = 2)
    )),
    xaxis = list(title = "Income Bracket", tickangle = -45),
    yaxis = list(title = "Savings Rate", tickformat = ".0%")
  )

```


Row {data-height=280}
-------------------------------------

### Savings Insight

```{r}

findings_savings <- tibble(
  Finding = c(
    "Brackets running deficit",
    "Worst deficit",
    "Breakeven bracket",
    "Best saving bracket",
    "Savings wealth gap"
  ),
  Insight = c(
    paste0(sum(bls_data$savings < 0), " of 8 brackets spend more than they earn"),
    paste0("$15-30k households run a ",
           percent(abs(bls_data$savings_rate[bls_data$bracket == "15_30k"]),
                   accuracy = 0.1), " deficit — spending 62 cents more than every dollar earned"),
    "$50-70k is where households first break even with a 0.3% savings rate",
    paste0("$200k+ households save ",
           percent(bls_data$savings_rate[bls_data$bracket == "200k_plus"],
                   accuracy = 0.1), " of income — ",
           dollar(bls_data$savings[bls_data$bracket == "200k_plus"], accuracy = 1), " per year"),
    paste0("Top earners save ",
           dollar(bls_data$savings[bls_data$bracket == "200k_plus"] -
                  bls_data$savings[bls_data$bracket == "70_100k"], accuracy = 1),
           " more per year than middle income households")
  )
)

knitr::kable(findings_savings, align = c("l", "l"))

```

Healthcare & Transport
=====================================

Row {data-height=500}
-------------------------------------

### Healthcare & Transport: Absolute Spend

```{r}

plot_ly(bls_data) %>%
  add_trace(x = ~income_bracket,
            y = ~healthcare,
            type = "bar",
            name = "Healthcare",
            marker = list(color = "#7b5ea7"),
            hovertemplate = "Bracket: %{x}<br>Healthcare: %{y:$,.0f}<extra></extra>") %>%
  add_trace(x = ~income_bracket,
            y = ~transport,
            type = "bar",
            name = "Transportation",
            marker = list(color = "#4ecdc4"),
            hovertemplate = "Bracket: %{x}<br>Transport: %{y:$,.0f}<extra></extra>") %>%
  layout(
    barmode = "group",
    xaxis = list(title = "Income Bracket", tickangle = -45),
    yaxis = list(title = "Annual Spend ($)", tickformat = "$,.0f"),
    legend = list(orientation = "h", y = -0.3)
  )

```

### As % of Income

```{r}

plot_ly(bls_data) %>%
  add_trace(x = ~income_bracket,
            y = ~healthcare / annual_income,
            type = "scatter",
            mode = "lines+markers",
            name = "Healthcare",
            line   = list(color = "#7b5ea7", width = 3),
            marker = list(color = "#7b5ea7", size = 8),
            hovertemplate = "Bracket: %{x}<br>Healthcare: %{y:.1%}<extra></extra>") %>%
  add_trace(x = ~income_bracket,
            y = ~transport / annual_income,
            type = "scatter",
            mode = "lines+markers",
            name = "Transport",
            line   = list(color = "#4ecdc4", width = 3),
            marker = list(color = "#4ecdc4", size = 8),
            hovertemplate = "Bracket: %{x}<br>Transport: %{y:.1%}<extra></extra>") %>%
  layout(
    xaxis = list(title = "Income Bracket", tickangle = -45),
    yaxis = list(title = "% of Income", tickformat = ".0%"),
    legend = list(orientation = "h", y = -0.3)
  )

```


Row {data-height=280}
-------------------------------------

### Healthcare & Transport Insight

```{r}

findings_ht <- tibble(
  Finding = c(
    "Transport dominates both categories",
    "Healthcare burden peaks low",
    "Transport gap top vs bottom",
    "Healthcare converges at top",
    "Combined burden"
  ),
  Insight = c(
    paste0("Transport exceeds healthcare spend across all brackets — ",
           dollar(bls_data$transport[bls_data$bracket == "200k_plus"], accuracy = 1),
           " vs ",
           dollar(bls_data$healthcare[bls_data$bracket == "200k_plus"], accuracy = 1),
           " for top earners"),
    paste0("$15-30k households spend ",
           percent(bls_data$healthcare[bls_data$bracket == "15_30k"] /
                   bls_data$annual_income[bls_data$bracket == "15_30k"], accuracy = 0.1),
           " of income on healthcare — highest of any bracket"),
    paste0("Top earners spend ",
           dollar(bls_data$transport[bls_data$bracket == "200k_plus"] -
                  bls_data$transport[bls_data$bracket == "15_30k"], accuracy = 1),
           " more on transport than the lowest bracket"),
    paste0("Healthcare as % of income falls to ",
           percent(bls_data$healthcare[bls_data$bracket == "200k_plus"] /
                   bls_data$annual_income[bls_data$bracket == "200k_plus"], accuracy = 0.1),
           " for $200k+ households — lowest of any bracket"),
    paste0("Combined healthcare and transport consume ",
           percent((bls_data$healthcare[bls_data$bracket == "15_30k"] +
                    bls_data$transport[bls_data$bracket == "15_30k"]) /
                   bls_data$annual_income[bls_data$bracket == "15_30k"], accuracy = 0.1),
           " of income for the lowest bracket")
  )
)

knitr::kable(findings_ht, align = c("l", "l"))

```
Key Findings
=====================================

Row {data-height=800}
-------------------------------------

### Summary of Analysis

```{r}

findings_master <- tibble(
  `#` = 1:7,
  Finding = c(
    "Lower income households run financial deficits",
    "Housing is the dominant burden across all brackets",
    "The cost burden breakeven is ~$70-100k",
    "Savings inequality is extreme",
    "Transport consistently exceeds healthcare spend",
    "Healthcare burden falls as income rises",
    "Real wages vs. real costs — a structural problem"
  ),
  Detail = c(
    "Brackets earning under $50k spend more than they earn — deficits range from 12% to 62% of income, based on 2024 BLS Consumer Expenditure Survey data",
    "Housing consumes 40%+ of total spend for all brackets and exceeds the HUD cost burden threshold of 30% of income for the bottom 5 of 8 brackets",
    "Only households earning $70k+ begin to meaningfully save — the $50-70k bracket barely breaks even at a 0.3% savings rate",
    "$200k+ households save $151,420/year vs. a deficit of $13,925 for $15-30k households — a gap of over $165,000 in annual financial position",
    "Transport spend exceeds healthcare across every income bracket — lowest earners spend 25% of income on transport alone",
    "Healthcare as % of income falls from 16.4% at the lowest bracket to 3.2% at the highest — a 5x difference in relative burden",
    "These patterns reflect structural cost pressures in housing and transport that disproportionately affect lower income households regardless of spending choices"
  )
)

knitr::kable(findings_master, align = c("c", "l", "l"))

```

---

**Data Source:** US Bureau of Labor Statistics, Consumer Expenditure Survey 2024.
Table 1203 — Income before taxes: Annual expenditure means, shares, standard errors,
and relative standard errors. Retrieved from [https://www.bls.gov/cex/tables.htm](https://www.bls.gov/cex/tables.htm)

**Note:** Savings calculated as income minus total expenditures. Negative savings
indicate expenditure exceeds reported income, which may reflect debt, asset drawdown,
or underreporting of income — a known limitation of the CE Survey methodology.